---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the ???quantified self??? movement ??? a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data for this assignment can be downloaded from the course web site:

Dataset: [Activity monitoring data] (https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)
The variables included in this dataset are:

steps: Number of steps taking in a 5-minute interval (missing values are coded as ????????)
date: The date on which the measurement was taken in YYYY-MM-DD format
interval: Identifier for the 5-minute interval in which measurement was taken
The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

The first step in the analysis is downloading and preprocessing of the raw data.

## Loading and preprocessing the data

The forementioned dataset is downloaded, unzipped and stored as variable "actdata". The variable "time" is also defined to be used later in the report.

```{r}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", destfile="activity.zip")
unzip("activity.zip")
actdata <- read.csv("./activity.csv")
time <- seq(from = as.POSIXct(x = "0:00", format = "%H:%M"), by = "5 mins",
            length.out = 288)
```

## What is mean total number of steps taken per day?

The first questions to be answered are: what are the total steps per day and put the results in a histogram that shows the distribution of the number of steps in the dataset.

```{r}
library(ggplot2)
daystepstotal <- aggregate(actdata$steps, 
                           list(actdata$date), 
                           sum)
daystepstotal$Group.1 <- as.Date(daystepstotal$Group.1,
                                 "%Y-%m-%d")
ggplot(daystepstotal,
       aes(x=x)) +
       geom_histogram(binwidth=1500,
                      position='identity') +
       ylab("Frequency") + xlab("# steps/day") +
  ggtitle("Histogram of total steps/day")

meansteps <- format(mean(daystepstotal$x,
                        na.rm = TRUE), scientific = FALSE)
mediansteps <- median(as.integer(daystepstotal$x),
                      na.rm = TRUE)

```

Next, the mean and median amount of total steps per day are calculated, which are `r meansteps` and `r mediansteps` respectively.

## What is the average daily activity pattern?

Another interesting bit of information to known is the distribution of the number of steps during the day. The next graph shows the distribution of the average number of steps during a day for each interval.

```{r}
daystepsmean <- aggregate(actdata$steps,
                          list(actdata$interval),
                          mean, na.rm = T)
plot(y = daystepsmean$x,
     x = time,
     type = "l",
     xlab = "Time (h:m)",
     ylab = "# steps per interval")

mostactive <- subset(daystepsmean,
                     x == max(daystepsmean$x))
mostactive <- format(strptime(sprintf("%04d", mostactive[,1]), format="%H%M"), format = "%H:%M")
```

You can see that at `r mostactive` , the subject is, on average, the most active.

## Imputing missing values

```{r}
NAs <- sum(is.na(actdata$steps))
perc_NAs <- format((NAs/nrow(actdata)*100), digits = 3)
```

Missing values can bias the results generated from a dataset. In this dataset there are `r NAs` missing values, which is `r perc_NAs`% of the total set.

To include the missing values, we replaced the missing values with the overall average of the specific interval. Below the histogram is shown similar to the previous histogram but now with the imputed dataset containing no missing values.

```{r}
actdata_impute <- actdata
actdata_impute$steps[is.na(actdata_impute$steps)] <-   as.integer(daystepsmean$x)
daystepstotal_impute <- aggregate(actdata_impute$steps,
                                  list(actdata_impute$date), sum)
daystepstotal_impute$Group.1 <- as.Date(daystepstotal_impute$Group.1,
                                        "%Y-%m-%d")
ggplot(daystepstotal_impute,
       aes(x=x)) +
       geom_histogram(binwidth=1500, position='identity') +
       ylab("Frequency") + xlab("# steps/day") +
       ggtitle("Histogram of total steps/day (imputed NA's)")

meansteps_impute<- format(mean(daystepstotal_impute$x,
                        na.rm = TRUE), scientific = FALSE)
mediansteps_impute<- median(as.integer(daystepstotal_impute$x), 
                            na.rm = TRUE)
```

Subsequently, the mean and median amount of total steps per day are calculated, which are `r meansteps_impute` and `r mediansteps_impute` respectively.

## Are there differences in activity patterns between weekdays and weekends?

Lastly, we analysed whether the activity patterns of weekdays and weekends are different. 

```{r}
actdata_impute$date1 <- weekdays(as.Date(actdata_impute$date))
actdata_impute_weekend <- subset(actdata_impute,
                                 date1 == "Saturday" | date1 == "Sunday")
weekendstepsmean <- aggregate(actdata_impute_weekend$steps,
                              list(actdata_impute_weekend$interval),
                              mean, 
                              na.rm = T)
actdata_impute_weekday <- subset(actdata_impute, 
                                 date1 == "Monday" | date1 == "Tuesday" |    date1 == "Wednesday" | date1 == "Thursday" | date1 == "Friday")
weekdaystepsmean <- aggregate(actdata_impute_weekday$steps,
                              list(actdata_impute_weekday$interval), 
                              mean, na.rm = T)

actdata_impute_weekday$date1 <- "weekday"
actdata_impute_weekend$date1 <- "weekend"
actdata_impute_necessary <- rbind(actdata_impute_weekday,
                                  actdata_impute_weekend)

plot <- par(mfrow=c(2,1), mar = c(2,4,1.5,1))
plot(y = weekdaystepsmean$x,
     x = time, 
     type = "l",
     ylab = "# steps per interval",
     main  = "Weekday")
plot(y = weekendstepsmean$x,
     x = time,
     type = "l",
     ylab = "# steps per interval",
     main = "Weekend")
```

Both patterns are similar, however small differences are evident. For example, in the weekends activity starts later and is less high and it continues later in the evening.

This concludes the analysis of the dataset as a part of the assignement in "Reproducable research"
